United States Impacts of Severe Weather
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Michael Nichols
Coursera - Reproducible Research: Course Project 2
## Synopsis
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. The analysis consists of tables, figures, and other summaries. 

The data collected is part of the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database, with events spreading from 1950 to 1995. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. The two main questions we'll be looking at within this dataset are listed below:

### Questions
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

### Data Processing
Load packages
```{r load_packages, message=FALSE, warning=FALSE}
library(plyr)
library(dplyr)
library(knitr)
library(ggplot2)
library(gridExtra)
```

Process:
1. set working directory
2. check if file exists & read data starting with original raw data file (repdata%2Fdata%2FStormData.csv.bz2) from the URL
```{r read_data}
#set WD
setwd("S:/Documents/R/Reproducible_Research2")

#download data from the url ("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2") & check if file exists
if(!file.exists("StormData.csv.bz2")) {
         url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
        download.file(url, destfile="StormData.bz2", method = "libcurl")
}

#load data 
data <- read.csv("StormData.bz2", stringsAsFactors = FALSE, strip.white=TRUE, header=TRUE)
```

Cleaning & filtering the dataset to display only the relevant columns within the dataset:

Relevant Data:
- EVTYPE: event type
- BGN_DATE: date the event occurred
- FATALITIES: count of fatalities
- INJURIES: count of injuries
- PROPDMG: damage to property in dollars
- PROPDMGEXP: exponent of PROPDMG (i.e. unit)
- CROPDMG: crop damage in dollars
- CROPDMGEXP: exponent of CROPDMG (i.e. unit)

Two new variables created:
- totalCPROPDMG
- totalCROPDMG

These will be numeric variables that find the total costs of multiplying PROPDMGEXP & CROPDMGEXP by the PROPDMG & CROPDMG fields

```{r clean_data, message=FALSE, warning=FALSE}
#retain only relevant data
relevantData <- data[ , c("EVTYPE", "BGN_DATE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
#display structure
str(relevantData)

#cleanse date field & shave off hour / minutes / seconds
relevantData$BGN_DATE <- as.Date(relevantData$BGN_DATE,"%m/%d/%Y")
str(relevantData$BGN_DATE)

#create clear costs of damages by concatenating propdmg & cropdmg with the exponent variables
#propdmg -- show exponents
sort(table(relevantData$PROPDMGEXP), decreasing = TRUE)
#cropdmg -- show exponents
sort(table(relevantData$CROPDMGEXP), decreasing = TRUE)

#K = thousands, M = millions, B = billions - explained on pg 12 - https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
#first create clean PROPDMGEXP & CROPDMGEXP var to create numeric exponents for billions, millions, thousands, hundreds & specifically listed exponents
cleanPROPDMGEXP <- mapvalues(relevantData$PROPDMGEXP,
                             c("K","M","", "B","m","+","0","5"),
                             c(1e3,1e6, 1, 1e9,1e6,  1,  1,1e5))

cleanCROPDMGEXP <- mapvalues(relevantData$CROPDMGEXP,
                             c("","M","K","m","B","?"),
                             c( 1,1e6,1e3,1e6,1e9,1))

#display new variable summary & confirm consistent conversions
sort(table(cleanPROPDMGEXP), decreasing = TRUE)
sort(table(cleanCROPDMGEXP), decreasing = TRUE)

#add variable to dataset based on PROPDMG & the new numeric exponentvariable
relevantData$totalPROPDMG <- as.numeric(cleanPROPDMGEXP) * relevantData$PROPDMG
relevantData$totalCROPDMG <- as.numeric(cleanCROPDMGEXP) * relevantData$CROPDMG

#remove new variables
remove(cleanPROPDMGEXP)
remove(cleanCROPDMGEXP)

#display summary of the costs
summary(relevantData$totalPROPDMG)
summary(relevantData$totalCROPDMG)

#show revised / clean dataset after removing old propdmg / cropdmg and exponent variables
cleanData <- relevantData[ , c("EVTYPE", "BGN_DATE", "FATALITIES", "INJURIES", "totalPROPDMG", "totalCROPDMG")]
str(cleanData)
head(cleanData)
```

### Summary of the data recorded
In order to gain a better understanding of when these data points were collected, the following plot will show how many events within the database were collected per year.

Display spread of when the data was recorded between 1950 & 1995.
```{r data_time_spread, message=FALSE, warning=FALSE}
#plot histogram to display the spread from what time this data was recorded
#strip year only from date field
cleanData$years <- as.numeric(format(cleanData$BGN_DATE, "%Y"))
summary(cleanData$years)

#create plot
p <- ggplot(cleanData, aes(x = years)) +
        geom_histogram(aes(fill = ..count..), binwidth = 1) + 
        ggtitle(label = "Dispersion of Data Collection", subtitle = "Spread of Events Recorded Between 1950 - 1995") +
        scale_x_continuous(name = "Years", breaks = seq(1950, 1995, 9), limits=c(1950,1995)) +
        scale_y_continuous(name = "Number of Events Recorded") +
        scale_fill_gradient("Events Recorded", low = "grey80", high = "dodgerblue3")

print(p)

#The chart clearly shows more data was gathered closer to 1995.
```

## Results 
Now that the data has been cleaned, the following information analyzes our two main questions.

### Question 1: Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```{r harmful_health_events}
#subset dataset to display only the fatalities & injuries for each event
fatalities <- aggregate(FATALITIES ~ EVTYPE, data = cleanData, FUN = sum)
injuries <- aggregate(INJURIES ~ EVTYPE, data = cleanData, FUN = sum)

#find top 10 most harmful for each category
#fatalities
#sort data frame in decreasing order and use head function to retain only top 10
fatalities <- fatalities[order(fatalities$FATALITIES, decreasing=TRUE),] 
fatalities <- head(fatalities, 6)

#injuries
#sort data frame in decreasing order and use head function to retain only top 10
injuries <- injuries[order(injuries$INJURIES, decreasing=TRUE),] 
injuries <- head(injuries, 6)

#create plot for fatalities
p1 <- ggplot(data = fatalities, aes(x = reorder(EVTYPE, -FATALITIES), y = FATALITIES, fill = EVTYPE)) +
  geom_bar(stat = "identity") +
  ggtitle(label = "Total Fatalities by Event Type", subtitle = "Top 6 Events from 1950-1995 in the Storm Database") +
  xlab("") + ylab("Fatalities Recorded") + 
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  geom_text(aes(label = FATALITIES))

#create plot for injuries
p2 <- ggplot(data = injuries, aes(x = reorder(EVTYPE, -INJURIES), y = INJURIES, fill = EVTYPE)) +
  geom_bar(stat = "identity") +
  ggtitle(label = "Total Injuries by Event Type", subtitle = "Top 6 Events from 1950-1995 in the Storm Database") +
  xlab("Event Type") + ylab("Injuries Recorded") + 
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  geom_text(aes(label = INJURIES)) 


# prep background for two plots - fatalities & injuries
grid.arrange(p1, p2, ncol = 1, heights = c(4,4))
```

### Question 2: Across the United States, which types of events have the greatest economic consequences?

```{r economic_consequences, message=FALSE, warning=FALSE}
#create new dataset to display aggregate damages by event
propdmg <- aggregate(totalPROPDMG ~ EVTYPE, data = cleanData, FUN = sum)
cropdmg <- aggregate(totalCROPDMG ~ EVTYPE, data = cleanData, FUN = sum)
economicDmg <- merge(propdmg, cropdmg, by.x = "EVTYPE")

#find top sources of prop and crop dmg
topPropDmg <- economicDmg[order(economicDmg$totalPROPDMG, decreasing=TRUE),]
topCropDmg <- economicDmg[order(economicDmg$totalCROPDMG, decreasing=TRUE),]

#filter to top 6 of each source and add label of damage type
topPropDmg <- head(topPropDmg, 6)
topPropDmg$Type <- "Property"
topPropDmg <- topPropDmg[,c("EVTYPE", "totalPROPDMG", "Type")]
topCropDmg <- head(topCropDmg, 6)
topCropDmg$Type <- "Crop"
topCropDmg <- topCropDmg[,c("EVTYPE", "totalCROPDMG", "Type")]

#merge top 6 sources of crop and prop dmg to new dataframe and create final variable called Damages which shows damages of top 6 sources of each
topEconomicDMG <- merge(topCropDmg, topPropDmg, all.x = TRUE, all.y = TRUE)
topEconomicDMG <- topEconomicDMG %>%
         rowwise() %>% 
         mutate(Damages = sum(totalCROPDMG, totalPROPDMG, na.rm = TRUE))


#create and print bar chart
p3 <- ggplot(data = topEconomicDMG, aes(x = reorder(EVTYPE, -Damages), y = Damages, fill = Type)) + 
         geom_bar(stat = "identity") + 
        ggtitle("Total Economic Damages by Event Type") + xlab("Event Type") + ylab("Damages (in US Dollars)") +
        theme(axis.text.x = element_text(angle=45, hjust=1))

print(p3)
```